---
editor_options: 
  chunk_output_type: console
---

# Onset times  

In this script, we will obtain the differences in times at which species vocalizes and plot it.  

## Install required libraries  

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(vegan)
library(ggplot2)
library(scico)
library(data.table)
library(extrafont)
library(showtext)
library(sf)
library(raster)
library(lubridate)
library(suncalc)
library(lutz)
library(stringr)
library(ggtext)
library(ggpubr)
library(hms)
library(lubridate)
library(ggrepel)
library(patchwork)
```

## Loading the data

```{r}
annotations <- read.csv ('results/data.csv')
sites <- read.csv("data/list-of-sites.csv")
species_codes <- read.csv("data/species-annotation-codes.csv")
eye_size_residuals <- read.csv ("results/eye_size_residuals.csv")
traits <- read.csv ('data/species-trait-dat.csv') 
territoriality <- read.csv ('data/territoriality-&-sociality-data.csv')
vegetation_structure <- read.csv ('data/2020-vegetation-data.csv')
```

## Modifying the species codes data for further analyses

```{r}
species_codes <- species_codes %>%
  rename (species = species_annotation_codes) #changing the name of the column so that it can be appended with the dataframe created in the downstream analyses
```

## Cleaning the territoriality data

```{r}
territoriality <- territoriality %>% 
   dplyr::select(Species, Territory) %>%
   rename (scientific_name = Species, territoriality = Territory) %>% #renaming it for easier downstream analyses
   mutate(scientific_name = recode(scientific_name, "Pycnonotus gularis" = "Rubigula gularis", "Ictinaetus malayensis" = "Ictinaetus malaiensis", "Dendrocopos nanus" = "Yungipicus nanus", "Megalaima malabarica" = "Psilopogon malabaricus", "Rhopocichla atriceps" = "Dumetia atriceps", "Chrysocolaptes lucidus" = "Chrysocolaptes guttacristatus", "Luscinia brunnea" = "Larvivora brunnea", "Parus aplonotus" = "Machlolophus aplonotus", "Turdoides striata" = "Argya striata", "Zoothera citrina" = "Geokichla citrina", "Turdoides subrufa" = "Argya subrufa", "Celeus brachyurus" = "Micropternus brachyurus", "Muscicapa ruficauda" = "Ficedula ruficauda", "Bubo nipalensis" = "Ketupa nipalensis", "Stigmatopelia chinensis" = "Spilopelia chinensis", "Acrocephalus aedon" = "Arundinax aedon", "Cyornis pallipes" = "Cyornis pallidipes", "Megalaima viridis" = "Psilopogon viridis", "Garrulax delesserti" = "Pterorhinus delesserti", "Acritillas indica" = "Iole indica")) #to update the names of these species as per our dataset
```

## Organizing the vegetation structure data- here, we consider foliage score which ranges from 1 to 8 indicating the presence/absence of foliage at different heights from the forest floor

```{r}
vert_strat <- vegetation_structure %>%
  dplyr::select (Site_ID, Foliage_score) %>% #retaining only selected columns
  rename (site_id = Site_ID) %>% #to rename the column
  mutate(across(everything(), ~str_replace_all(., "_", ""))) %>%
  group_by (site_id) %>%
  mutate (Foliage_score = as.numeric (Foliage_score)) %>% 
  summarise(vert_strat = mean(Foliage_score)) #this gives mean number of forest layers
```

## Extracting the nautical times for each site and date

```{r}
annotations <- annotations %>%
  filter (!species == 'UK') %>% # removing all unidentified annotations
  rename (site_id = site_code) %>% #changing the name of the column as per the site data
  left_join (., sites, by = 'site_id') %>% # joining the two dataframes
  dplyr::select(!c(notes, forest_name, Year.Active, No..saplings.planted, No..species.planted, Planted.area..ha., Distance.to.continuous.forest..m., Remnant.area..ha., Notes,)) %>% # removing unwanted columns
  rename (lat = latitude, lon = longitude) %>% # renaming the columns
  ungroup ()

annotations$date <- dmy (annotations$date) #reading the dates in dmy format

nautical_dawn <- getSunlightTimes(data = annotations, 
  keep = c(
    "sunrise", 
    "nauticalDawn"), #keeping sunrise and nautical dawn times
  tz = "Asia/Kolkata"
) %>% 
  distinct(.) 

nautical_dawn$nauticalDawn <- as_hms(nautical_dawn$nauticalDawn) #reading the times in hms format

annotations <- annotations %>%
  left_join (., nautical_dawn, by = c('lat','lon', 'date')) #now, we have a dataframe with species, their onset times, dates, nautical dawn times, and sunrise times corresponding to each annotation at each site for both the seasons. 
```

## Filtering species with adequate observations

```{r}
sp <- annotations %>%
  group_by (species) %>%
  summarize (n= n()) %>%
  filter (!n<10) %>%
  pull (species) #this helps us only retain those species for which we have atleast 10 observations, i.e. 89 species.
```

## Calculating relative onset times of each observation, and adding species-specific information to this data

```{r}
consolidated_data <- annotations %>%
  filter (species %in% sp) %>% #filtering the selected species
  dplyr::select (site_id, filename, date, begin_clock_time, species, nauticalDawn, sunrise, peak_freq, season) %>%

#adding data on site-wise vertifical stratification 
  left_join (., vert_strat, by = 'site_id') %>%
  
#calculating relative onset times as the differences between timing of first vocalization of each species and the nautical dawn at corresponding site, and converting it into minutes by dividing the amount in seconds by 60, before summarizing it as a median
 mutate (begin_clock_time_hms = as.POSIXct(begin_clock_time, format = "%H:%M:%S")) %>% #converting to a Posixct object 
  mutate (begin_clock_time_hms = as_hms (begin_clock_time_hms)) %>% #retaining only hours and mins information
  mutate(relative_onset_times = as.numeric (begin_clock_time_hms-nauticalDawn), relative_onset_times = relative_onset_times / 60) %>% 
  group_by (species) %>%
 
  #joining the species annotation codes dataset
  inner_join (., species_codes, by = 'species') %>% 
  mutate(scientific_name = recode(scientific_name, "Iole indica" = "Acritillas indica")) %>% #updating the scientific name of yellow-browed bulbul

#adding relative eye size estimates/residuals of the species
  left_join (., eye_size_residuals, by = c('scientific_name', 'common_name')) %>% 
  na.omit () %>% 
  mutate(scientific_name = recode(scientific_name,"Acritillas indica" =  "Iole indica")) %>% #changing the scientific name again 
  
#adding territoriality information of the species
  left_join (., territoriality, by = 'scientific_name') %>% #joining it with the onset times, i.e. outcome dataframe
  mutate(territoriality = case_when(territoriality %in% "1" ~ "Non-territorial", territoriality %in% "2" ~ "Weakly territorial", territoriality %in% "3" ~ "Highly territorial")) %>%
  
#adding trait information of foraging height, body mass, and trophic niche of the species
  left_join (., traits, by = c('scientific_name', 'common_name')) %>%
  dplyr::select (site_id, date, filename, species, begin_clock_time_hms, nauticalDawn, sunrise, peak_freq, vert_strat, relative_onset_times, scientific_name, common_name, ebird_residuals, foraging_habit, territoriality, mass, trophic_niche, season)

#This dataframes gives us data for 74 species (and it doesn't include any raptors).

write.csv (consolidated_data, 'results/consolidated_data.csv')
```

## Creating a dataframe with median, 1st and 3rd quantiles (for reference) for colouring the species accordingly 

```{r}
median_onset_times <- consolidated_data %>%
  group_by (species, common_name) %>%
  summarise (median_relative_onset_time = median(relative_onset_times), q1 = quantile(relative_onset_times, 0.25), q3 = quantile(relative_onset_times, 0.75)) %>%
  #giving different colours to indicate the differences in relative onset times
  mutate (colour = case_when (median_relative_onset_time > 0 & median_relative_onset_time <= 20 ~ '#FFB000',
                              median_relative_onset_time > 20 & median_relative_onset_time <= 40 ~ '#009E73',
                              median_relative_onset_time > 40 & median_relative_onset_time <= 60 ~ '#000000',
                              median_relative_onset_time > 60 & median_relative_onset_time <= 80 ~ '#FE6100',
                              median_relative_onset_time > 80 ~ '#785EF0')) %>%
  arrange(median_relative_onset_time) %>%
  mutate(common_name_styled = paste0("<span style='color:", colour, "'>", common_name, "</span>"), #creating the HTML-styled labels for coloring
     common_name_ordered = fct_reorder(common_name, median_relative_onset_time))  #create an ordering factor
```

## Adding the species colour codes information to the consolidated_data

```{r}
consolidated_data <- consolidated_data %>% 
  left_join (median_onset_times,., by = c('species', 'common_name')) %>% #joining the dataframes so that we can add the colour codes info
  na.omit () %>%
  arrange(median_relative_onset_time) 
```

## i) Plotting the data- for both seasons

```{r}
rel_onset_times_boxplot <- ggplot(consolidated_data, aes(x = fct_reorder (common_name,median_relative_onset_time), y = relative_onset_times, color = colour)) +
  geom_boxplot (outlier.shape = NA) +
  geom_jitter (shape = 20, size = 1.2, alpha = 0.5) +
  coord_flip() +
  theme_bw() +
  scale_x_discrete(labels = setNames(consolidated_data$common_name_styled, consolidated_data$common_name_ordered)) +
  scale_color_identity(guide = "legend", "Difference between onset times and nautical dawn times", labels = c('between 40 to 60 mins', 'between 20 to 40 mins', 'more than 80 mins', 'between 60 to 80 mins', 'between 0 to 20 mins')) + 
  labs(y = 'Relative onset times (minutes)', x = 'Species common name') +
  theme(
    axis.text.y = element_markdown(family = 'Century Gothic', size = 14),
    axis.text.x = element_text(family = 'Century Gothic', size = 14), 
    text = element_text(family = 'Century Gothic'),
    axis.title = element_text(family = 'Century Gothic', size = 20, face = 'bold'),
    legend.text = element_text(family = 'Century Gothic', size = 14),
    legend.title = element_text(family = 'Century Gothic', size = 20)
  )

# inspect the visualization
rel_onset_times_boxplot

# save the plot
ggsave (rel_onset_times_boxplot,
  filename = "figs/rel_onset_times.png",  width    = 16,
  height   = 18,
  units    = "in",
  dpi      = 300
)
dev.off()
```

## ii) Plotting the data- for each season separately

## Making separate dataframes for each season

```{r}
summer_consolidated_data <- consolidated_data %>%
  filter (season %in% 'summer') 
winter_consolidated_data <- consolidated_data %>%
  filter (season %in% 'winter')
```

## Extracting species common across both seasons

```{r}
common_sp <- intersect (summer_consolidated_data$scientific_name, winter_consolidated_data$scientific_name) #finding the species common across both seasons, i.e. 72 species
```

## Now, filtering the species common across both seasons

```{r}
summer_consolidated_data <- summer_consolidated_data %>%
  filter (scientific_name %in% common_sp) 
winter_consolidated_data <- winter_consolidated_data %>%
  filter (scientific_name %in% common_sp)
```

## Creating dataframes for each season with median, 1st and 3rd quantiles (for reference) for colouring the species accordingly

```{r}
summer_median_onset_times <- summer_consolidated_data %>%
  group_by (species, common_name) %>%
  summarise (median_relative_onset_time = median(relative_onset_times), q1 = quantile(relative_onset_times, 0.25), q3 = quantile(relative_onset_times, 0.75)) %>%
  mutate (colour = case_when (median_relative_onset_time > 0 & median_relative_onset_time <= 20 ~ '#FFB000',
                              median_relative_onset_time > 20 & median_relative_onset_time <= 40 ~ '#009E73',
                              median_relative_onset_time > 40 & median_relative_onset_time <= 60 ~ '#000000',
                              median_relative_onset_time > 60 & median_relative_onset_time <= 80 ~ '#FE6100',
                              median_relative_onset_time > 80 ~ '#785EF0')) %>%
  arrange(median_relative_onset_time) %>%
  mutate(common_name_styled = paste0("<span style='color:", colour, "'>", common_name, "</span>"), 
     common_name_ordered = fct_reorder(common_name, median_relative_onset_time)) 

winter_median_onset_times <- winter_consolidated_data %>%
  group_by (species, common_name) %>%
  summarise (median_relative_onset_time = median(relative_onset_times), q1 = quantile(relative_onset_times, 0.25), q3 = quantile(relative_onset_times, 0.75)) %>%
  mutate (colour = case_when (median_relative_onset_time > 0 & median_relative_onset_time <= 20 ~ '#FFB000',
                              median_relative_onset_time > 20 & median_relative_onset_time <= 40 ~ '#009E73',
                              median_relative_onset_time > 40 & median_relative_onset_time <= 60 ~ '#000000',
                              median_relative_onset_time > 60 & median_relative_onset_time <= 80 ~ '#FE6100',
                              median_relative_onset_time > 80 ~ '#785EF0')) %>%
  arrange(median_relative_onset_time) %>%
  mutate(common_name_styled = paste0("<span style='color:", colour, "'>", common_name, "</span>"), 
     common_name_ordered = fct_reorder(common_name, median_relative_onset_time))
```

## Plotting the data

```{r}
sum_rel_onset_times_boxplot <- ggplot(summer_consolidated_data, aes(x = fct_reorder (common_name,median_relative_onset_time), y = relative_onset_times, color = colour)) +
  geom_boxplot (outlier.shape = NA) +
  geom_jitter (shape = 20, size = 1.2, alpha = 0.5) +
  coord_flip() +
  theme_bw() +
  scale_x_discrete(labels = setNames(summer_consolidated_data$common_name_styled, summer_consolidated_data$common_name_ordered)) +
  scale_color_identity(guide = "legend", "Difference between onset times and nautical dawn times", labels = c('between 40 to 60 mins', 'between 20 to 40 mins', 'more than 80 mins', 'between 60 to 80 mins', 'between 0 to 20 mins')) + 
  labs(y = 'Relative onset times (minutes)', x = 'Species common name') +
  theme(
    axis.text.y = element_markdown(family = 'Century Gothic', size = 14),
    axis.text.x = element_text(family = 'Century Gothic', size = 14), 
    text = element_text(family = 'Century Gothic'),
    axis.title = element_text(family = 'Century Gothic', size = 20, face = 'bold'),
    legend.text = element_text(family = 'Century Gothic', size = 14),
    legend.title = element_text(family = 'Century Gothic', size = 20)
  )

# inspect the visualization
sum_rel_onset_times_boxplot

# save the plot
ggsave (sum_rel_onset_times_boxplot,
  filename = "figs/sum_rel_onset_times.png",  width    = 16,
  height   = 18,
  units    = "in",
  dpi      = 300
)
dev.off()

win_rel_onset_times_boxplot <- ggplot(winter_consolidated_data, aes(x = fct_reorder (common_name,median_relative_onset_time), y = relative_onset_times, color = colour)) +
  geom_boxplot (outlier.shape = NA) +
  geom_jitter (shape = 20, size = 1.2, alpha = 0.5) +
  coord_flip() +
  theme_bw() +
  scale_x_discrete(labels = setNames(winter_consolidated_data$common_name_styled, winter_consolidated_data$common_name_ordered)) +
  scale_color_identity(guide = "legend", "Difference between onset times and nautical dawn times", labels = c('between 40 to 60 mins', 'between 20 to 40 mins', 'more than 80 mins', 'between 60 to 80 mins', 'between 0 to 20 mins')) + 
  labs(y = 'Relative onset times (minutes)', x = 'Species common name') +
  theme(
    axis.text.y = element_markdown(family = 'Century Gothic', size = 14),
    axis.text.x = element_text(family = 'Century Gothic', size = 14), 
    text = element_text(family = 'Century Gothic'),
    axis.title = element_text(family = 'Century Gothic', size = 20, face = 'bold'),
    legend.text = element_text(family = 'Century Gothic', size = 14),
    legend.title = element_text(family = 'Century Gothic', size = 20)
  )

# inspect the visualization
win_rel_onset_times_boxplot

# save the plot
ggsave (win_rel_onset_times_boxplot,
  filename = "figs/win_rel_onset_times.png",  width    = 16,
  height   = 18,
  units    = "in",
  dpi      = 300
)
dev.off()

```


```{r}
#making separate dataframes for each season
summer_consolidated_data <- consolidated_data %>%
  filter (season %in% 'summer') 
winter_consolidated_data <- consolidated_data %>%
  filter (season %in% 'winter')
 
common_sp <- intersect (summer_consolidated_data$scientific_name, winter_consolidated_data$scientific_name) #finding the species common across both seasons, i.e. 72 species

#filtering the common species
summer_consolidated_data <- summer_consolidated_data %>%
  filter (scientific_name %in% common_sp) 
winter_consolidated_data <- winter_consolidated_data %>%
  filter (scientific_name %in% common_sp) 

sum_rel_onset_times_boxplot <- ggplot(summer_consolidated_data, aes(x = fct_reorder (common_name,median_relative_onset_time), y = relative_onset_times, color = colour)) +
  geom_boxplot (outlier.shape = NA) +
  geom_jitter (shape = 20, size = 1.2, alpha = 0.5) +
  coord_flip() +
  theme_bw() +
  scale_x_discrete(labels = setNames(summer_consolidated_data$common_name_styled, summer_consolidated_data$common_name_ordered)) +
  scale_color_identity(guide = "legend", "Difference between onset times and nautical dawn times", labels = c('between 40 to 60 mins', 'between 20 to 40 mins', 'more than 80 mins', 'between 60 to 80 mins', 'between 0 to 20 mins')) + 
  labs(y = 'Relative onset times (minutes)', x = 'Species common name') +
  theme(
    axis.text.y = element_markdown(family = 'Century Gothic', size = 14),
    axis.text.x = element_text(family = 'Century Gothic', size = 14), 
    text = element_text(family = 'Century Gothic'),
    axis.title = element_text(family = 'Century Gothic', size = 20, face = 'bold'),
    legend.text = element_text(family = 'Century Gothic', size = 14),
    legend.title = element_text(family = 'Century Gothic', size = 20)
  )

# inspect the visualization
sum_rel_onset_times_boxplot

# save the plot
ggsave (sum_rel_onset_times_boxplot,
  filename = "figs/sum_rel_onset_times.png",  width    = 16,
  height   = 18,
  units    = "in",
  dpi      = 300
)
dev.off()

win_rel_onset_times_boxplot <- ggplot(winter_consolidated_data, aes(x = fct_reorder (common_name,median_relative_onset_time), y = relative_onset_times, color = colour)) +
  geom_boxplot (outlier.shape = NA) +
  geom_jitter (shape = 20, size = 1.2, alpha = 0.5) +
  coord_flip() +
  theme_bw() +
  scale_x_discrete(labels = setNames(winter_consolidated_data$common_name_styled, winter_consolidated_data$common_name_ordered)) +
  scale_color_identity(guide = "legend", "Difference between onset times and nautical dawn times", labels = c('between 40 to 60 mins', 'between 20 to 40 mins', 'more than 80 mins', 'between 60 to 80 mins', 'between 0 to 20 mins')) + 
  labs(y = 'Relative onset times (minutes)', x = 'Species common name') +
  theme(
    axis.text.y = element_markdown(family = 'Century Gothic', size = 14),
    axis.text.x = element_text(family = 'Century Gothic', size = 14), 
    text = element_text(family = 'Century Gothic'),
    axis.title = element_text(family = 'Century Gothic', size = 20, face = 'bold'),
    legend.text = element_text(family = 'Century Gothic', size = 14),
    legend.title = element_text(family = 'Century Gothic', size = 20)
  )

# inspect the visualization
win_rel_onset_times_boxplot

# save the plot
ggsave (win_rel_onset_times_boxplot,
  filename = "figs/win_rel_onset_times.png",  width    = 16,
  height   = 18,
  units    = "in",
  dpi      = 300
)
dev.off()
```

