---
title: "04_scm_pgls_analyses"
output: html_document
---

# SCM_PGLS_analysis

In this script, we will examine which of the causal variables (peak frequency, vegetation structure, trophic niche, foraging height, eye size, territoriality) best explain the differences in onset times using the structural causal model framework and the analysis method of phylogenetic generalized least squares regression. We use the phylogenetic tree from McTavish et al. 2025. 

## Install required libraries

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(vegan)
library(ggplot2)
library(scico)
library(data.table)
library(extrafont)
library(sf)
library(raster)
library(lubridate)
library(suncalc)
library(lutz)
library(stringr)
library(ggtext)
library(ggpubr)
library(hms)
library(clootl)
library(ape)
library(nlme)
library(geiger)
library(phytools)
library(phangorn)
library(phyr)
library(psych)
```

## Reading the required data

```{r}
consolidated_data <- read.csv ('results/consolidated_data.csv')
```

## Cleaning the data

```{r}
consolidated_data <- consolidated_data %>%
  filter(trophic_niche != "Aquatic predator") %>%
  filter(trophic_niche != "Granivore") %>%
  filter(trophic_niche != "Nectarivore") %>%
  filter(trophic_niche != "Vertivore") %>%
  filter (foraging_habit != "AER") %>%
  filter (foraging_habit != "AQU") #removing trophic niches and foraging habit categories that are poorly represented (Aquatic predator- 1 sp; Granivore- 1 sp; Nectarivore- 2 sp; Vertivore- 3 sp; AQU- 1 sp; AER- 1 sp)
  
#This leaves us with 67 species for downstream analyses.
```

## Extracting the list of species (scientific names) for phylogeny

```{r}
species <- consolidated_data %>%
  distinct (scientific_name) %>%
  #unique (consolidated_data$scientific_name)
  mutate(scientific_name = recode(scientific_name, "Accipiter badius" = "Tachyspiza badia"), 
        scientific_name = recode (scientific_name, "Iole indica" = "Acritillas indica"))

species <- as.character (species$scientific_name) #converting it into character vector
```

## Extracting a phylogenetic tree of the 67 species

```{r}
tree <- extractTree (species = species, label_type= "scientific", taxonomy_year= "2025")
plot (tree)
```

## Changing the names of two species to match the phylogenetic tree

```{r}
consolidated_data <- consolidated_data %>%
  mutate(scientific_name = recode(scientific_name, "Accipiter badius" = "Tachyspiza badia"),   scientific_name = recode (scientific_name, "Iole indica" = "Acritillas indica")) %>%
  mutate(across(scientific_name, str_replace, " ", "_"))
```

## Checking species in the tree and the data

```{r}
setdiff (consolidated_data$scientific_name, tree$tip.label) #to check if all species in the data are on the tree: result 'character (0)'

setdiff (tree$tip.label, consolidated_data$scientific_name) #to check if all species on the tree are in the data: result 'character (0)'
```

## Testing for correlations among the variables

```{r}
for_corr <- consolidated_data %>%
  dplyr::select (scientific_name, relative_onset_times, ebird_residuals, mass, trophic_niche, foraging_habit, peak_freq, territoriality, vert_strat) %>%
  mutate (relative_onset_time_z = scale (relative_onset_times), ebird_residuals_z = scale (ebird_residuals), pf_z = scale (peak_freq), vert_strat_z = scale (vert_strat)) %>% #scaling the numerical variables
  dplyr::select (scientific_name, relative_onset_time_z, ebird_residuals_z, mass, trophic_niche, foraging_habit, pf_z, territoriality, vert_strat_z) #only keeping the scaled variables

pairs.panels(for_corr[, c(2:9)]) #none of the variables are significantly correlated with each other, so we retain all of them
```

## PGLS Analysis

```{r}
consolidated_data <- consolidated_data  %>%
  mutate (ebird_residuals_z = scale (ebird_residuals), peak_freq_z = scale (peak_freq), vert_strat_z = scale (vert_strat), mass_z = scale (mass))

#Relative onset times ~ trophic niche (adjustment sets: Eye size, Foraging height, Vegetation structure)
consolidated_data$trophic_niche <- relevel(factor(consolidated_data$trophic_niche), ref = "Frugivore")
consolidated_data$foraging_habit <- relevel(factor(consolidated_data$foraging_habit), ref = "CAN")

# throwing an error currently because your species in the tree are mismatched
# once that is fixed, this code should run
model1 <- consolidated_data %>%
  gls(relative_onset_times ~ trophic_niche +  
                                   ebird_residuals +  
                                   foraging_habit +
                                   vert_strat,
    data = .,
    correlation = corBrownian(1, tree, form = ~scientific_name))

summary(model1) 
confint (model1) #only run this model

#Relative onset times ~ territoriality (adjustment sets: none)
model2 <- consolidated_data %>%
  gls(median_relative_onset_time ~ relevel(factor(territoriality), ref = "Non-territorial"),
    data = .,
    correlation = corBrownian(1, tree, form = ~scientific_name))

summary(model2)
confint (model2)

#Relative onset times ~ foraging height (adjustment sets: Eye Size, Vegetation structure)
model3 <- consolidated_data %>%
  gls(median_relative_onset_time ~ relevel(factor(foraging_habit), ref = "GRD") + 
                                   ebird_residuals_z +
                                   mean_vert_strat_z,
    data = .,
    correlation = corBrownian(1, tree, form = ~scientific_name))

summary(model3)
confint (model3)

#Relative onset times ~ eye size (adjustment sets: Foraging Height, Vegetation Structure)
model4 <- consolidated_data %>%
  gls(median_relative_onset_time ~ ebird_residuals_z + 
                                   relevel(factor(foraging_habit), ref = "GRD") + 
                                   mean_vert_strat_z,
    data = .,
    correlation = corBrownian(1, tree, form = ~scientific_name))

summary(model4)
confint (model4)

#plotting the results of the model since eye size shows a marginally significant cause-effect relationship with onset times (p = 0.0896)
font_add('fa-solid', 'data/fonts/Font Awesome 7 Free-Solid-900.otf') #loading solid font for fontawesome icon of eyes on the plot

eye_size_model_plot <- ggplot(consolidated_data, aes(x=ebird_residuals_z, y=median_relative_onset_time)) +
  geom_point(shape = 20, size = 5) +
  geom_abline(intercept = coef(model4)[1],
              slope = coef(model4)[2]) +
 labs(y = 'Median relative onset times (minutes)', x = 'Residual eye size') +
  theme_bw() +
  theme(axis.text.y = element_markdown(family = 'Century Gothic', size = 12),
    axis.text.x = element_text(family = 'Century Gothic', size = 12), 
    text = element_text(family = 'Century Gothic'), 
    axis.title = element_text(family = 'Century Gothic', size = 18, face = 'bold')) +
  annotate("text", x = 1,
     y = 125,
     label = paste ("estimate = -7.124; std error = 4.129; p = 0.0896"),
    color = "black",
     family = "Century Gothic", size = 10) +
  geom_richtext(x = -2.5, y = 48, label = "<span style='font-family:fa-solid;'>&#xf06e;</span>", size = 15, label.colour = NA, col = '#FE6100') +
  geom_richtext(x = 2, y = 13, label = "<span style='font-family:fa-solid;'>&#xf06e;</span>", size = 30, label.colour = NA, col = '#FE6100')

# inspect the visualization
eye_size_model_plot

# save the plot
ggsave (eye_size_model_plot,
  filename = "figs/eye_size_model_plot.png",  width    = 8,
  height   = 6,
  units    = "in",
  dpi      = 300
)
dev.off()
  
#Relative onset times ~ vegetation structure (adjustment sets: Eye Size, Foraging Height, Peak Frequency)
model5 <- consolidated_data %>%
  gls(median_relative_onset_time ~ mean_vert_strat_z +
                                   ebird_residuals_z + 
                                   relevel(factor(foraging_habit), ref = "GRD") +
                                   median_pf_z,
    data = .,
    correlation = corBrownian(1, tree, form = ~scientific_name))

summary(model5)
confint (model5)

#Relative onset times ~ peak frequency (adjustment sets: Vegetation Structure)
model6 <- consolidated_data %>%
  gls(median_relative_onset_time ~ median_pf_z + 
                                   mean_vert_strat_z,
    data = .,
    correlation = corBrownian(1, tree, form = ~scientific_name))

summary(model6)
confint (model6)
```





